---
title: "Final Project:"
subtitle: "Predicting _____ Using ____ Data"
author: "Names in Alphabetical order by surname"
date: today
format:
  html:
    code-fold: show
    code-summary: "Show Code"
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Introduction

\[Brief write up explaining what we chose, why, and why it's important\]

We are using eviction filing data from the Eviction Lab to create a predictive model to predict which census tracts in Philadelphia are most at risk of eviction filings. The goal of this predictive model is to help the city target housing stability interventions more effectively.

We use \_\_\_\_\_\_ features....

# Research Goals

**Can we predict in which census tracts evictions are likely to occur based on where they occurred in the past?**

# Set Up

```{r setup}
#| message: false
#| warning: false

library(tidyr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(viridis)
library(knitr)
library(kableExtra)
library(patchwork)
library(scales)
library(here)
library(stringr)
library(MASS)
options(scipen = 999)

```

# Import and Prep Data

For this report, we are using eviction filing data spanning 2020-April 2026. This data is made available online through thge Eviction Lab and was accessed through this link: https://evictionlab.org/eviction-tracking/philadelphia-pa/

## Load Eviction Data

First, we are loading in our eviction filing data - document other data loaded - document clenaing steps

```{r load_evictions_data}
#| message: false
#| warning: false

eviction_data <- read.csv("data/philadelphia_weekly_2020_2021.csv")

check_one_tract <- eviction_data %>%
  filter(GEOID == "42101027200")

# I changed the data-set to get the census tracts and to make sure all the evictions were located in Philadelphia 
eviction_data <- eviction_data %>% 
  mutate(
    state  = substr(GEOID, 1, 2),
    county = substr(GEOID, 3, 5),
    tract  = paste0("Census Tract ", as.numeric(substr(GEOID, 6, 11)) / 100)
  ) %>%  
  filter(state == "42", county == "101")

#Added the month and year as seperate columns 
eviction_data<- eviction_data %>% 
  mutate(year = year(week_date),
         month= month(week_date))

#I'm adding a season column 
eviction_data<- eviction_data %>% 
 mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "winter",
    month %in% c(3, 4, 5) ~ "spring",
    month %in% c(6, 7, 8) ~ "summer",
    TRUE ~ "fall"
  ))

monthly_filings <- read.csv("data/philadelphia_barchart.csv")

filings_monthly <- read.csv("data/philadelphia_monthly_2020_2021.csv")

#This breaks down the filings for each month, year and census tract 
eviction_data_1<- eviction_data %>% 
  group_by(GEOID, tract, year, month) %>% 
  summarise(n= sum(filings_2020))

eviction_data_1<- eviction_data_1 %>% 
  rename(GEOID10= "GEOID")



#Philadelphia census tract boundaries 

census_tracts<- st_read("data/census_tracts.shp")

if (st_crs(census_tracts)$epsg != 2272) {
  census_tracts <- st_transform(census_tracts, 2272)
}

```

##Exploratory Data Analysis

```{r}
#| message: false
#| warning: false

##Map all years 

##Plot all years for a single tract
check_one_tract %>%
  mutate(year = lubridate::year(week_date)) %>%
  group_by(year, week) %>%
  summarize(
    filings = sum(filings_2020, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = filings, color = factor(year))) +
  geom_line() +
  labs(title = "Weekly Eviction Filings by Year")

#Code to plot weekly filings in 2020 for one tract

check_one_tract %>%
  filter(lubridate::year(week_date) == 2020) %>%
  group_by(week) %>%
  summarize(
    filings_2020 = sum(filings_2020, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = week, y = filings_2020)) +
  geom_line() +
  labs(title = "Weekly Eviction Filings (2020 only)")

##
# Total citywide filings per week
weekly_plot <- eviction_data %>%
  mutate(year = lubridate::year(week_date)) %>%
  group_by(year, week) %>%
  summarize(
    total_filings = sum(filings_2020, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = total_filings, color = factor(year))) +
  geom_line(linewidth = 1) +
  labs(
    title = "Total Weekly Eviction Filings (Philadelphia Citywide)",
    x = "Week",
    y = "Total Filings",
    color = "Year"
  ) +
  theme_minimal()

weekly_plot

##Histogram
ggplot(eviction_data, aes(x = filings_2020)) +
  geom_histogram(bins = 50)

##seasonal patterns 
seasonal_patterns <- eviction_data %>%
  group_by(season) %>%
  summarize(
    avg_filings = mean(filings_2020, na.rm = TRUE),
    .groups = "drop"
  )


ggplot(seasonal_patterns, aes(x = season, y = avg_filings, group = 1)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Seasonal Eviction Patterns",
    x = "Season",
    y = "Average Filings"
  )


##Map
eviction_tract <- eviction_data %>%
  group_by(GEOID) %>%
  summarize(
    avg_filings = mean(filings_2020, na.rm = TRUE),
    total_filings = sum(filings_2020, na.rm = TRUE),
    persistence = mean(filings_2020 > 0, na.rm = TRUE),
    .groups = "drop"
  )



census_tracts <- census_tracts %>%
  rename(GEOID = GEOID10)

philly_map <- census_tracts %>%
  left_join(eviction_tract, by = "GEOID") 

library(ggspatial)
eviction_map <- ggplot(philly_map) +
  geom_sf(aes(fill = avg_filings), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c() +
  
  # --- North arrow ---
  annotation_north_arrow(
    location = "br",   # top right
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # --- Scale bar ---
  annotation_scale(
    location = "bl",   # bottom left
    width_hint = 0.3
  ) +
  
  labs(
    title = "Average Eviction Filings by Census Tract (Philadelphia)",
    fill = "Avg filings"
  ) +
  theme_void()

eviction_map

```

#Eviction Proxy Variables

## Bike Lanes

```{r}
#| message: false
#| warning: false


#The idea here is we may see a pattern where eviction notices are concentrated in census tracts where bike infrastructure is strong in the city 

bike_data<- st_read("data/Bike_Network.shp")

bike_data <- st_transform(bike_data, st_crs(census_tracts))

#This gives me all the census tracts where bike lanes are located 
bike_census<- st_join(bike_data, census_tracts, join = st_intersects) %>% 
  group_by(NAMELSAD10) %>% 
  tally()

#This gives me all the census tracts where evictions occur 
eviction_tracts_unique <- eviction_data %>% 
  distinct(tract)

#So now, I want to make sure that my bike_census contains all the census tracts that are in the eviction filings dataset and not ones that are not. (I think this is the right thing to do for consistency)

setequal(bike_census$NAMELSAD10, eviction_tracts_unique$tract)

#Values in A that are not in B so census tracts that I have in the bike_census data set that is not in the eviction filings data_set 
setdiff(bike_census$NAMELSAD10, eviction_tracts_unique$tract)

bad_tracts <- setdiff(bike_census$NAMELSAD10, eviction_tracts_unique$tract)

bike_census <- bike_census %>%
  filter(!NAMELSAD10 %in% bad_tracts)

bike_census<- bike_census %>% 
  rename(tract= NAMELSAD10)

```

## Landmarks

```{r}
#| message: false
#| warning: false


landmarks<- st_read("data/Landmark_Points.shp")

landmarks_subtype<- landmarks %>% 
  group_by(SUBTYPE) %>% 
  tally()

landmarks<- st_transform(landmarks, st_crs(census_tracts))

landmarks<- st_join(landmarks, census_tracts)

landmarks<- landmarks %>%  
  dplyr::select(NAME, TYPE, SUBTYPE, NAMELSAD10) %>% 
  st_drop_geometry() %>% 
  group_by(NAMELSAD10) %>% 
   summarise(landmarks = n()) %>%
  rename(tract= NAMELSAD10)

```

## Census Variables

```{r}
#| message: false
#| warning: false


census<- get_acs( 
  geography = "tract",
  variables = c(
    "B19013_001",  # Median household income
    "B08301_010E",  # Transit commuters
    "B01002_001E" #Median age 
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide",
  progress_bar = FALSE
)%>%
  rename(
    Med_Inc = B19013_001E,
    Transit_Commuters = B08301_010E,
    Median_age= B01002_001E
  )

census <- st_transform(census, 2272)

# Verify
st_crs(census)$epsg

```

## Mobility

```{r}
#| message: false
#| warning: false

philly_mobility <- get_acs(
  geography = "tract",
  variables = c(
    "B07003_001",
    "B07003_004",
    "B07003_007",
    "B07003_010",
    "B07003_013",
    "B07003_016"
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide"
)

# Rename
philly_mobility <- philly_mobility %>%
  rename(
    total_pop_1yr = B07003_001E,
    same_house = B07003_004E,
    moved_same_county = B07003_007E,
    moved_diff_county_same_state = B07003_010E,
    moved_diff_state = B07003_013E,
    moved_abroad = B07003_016E
  )

philly_mobility <- philly_mobility %>%
  mutate(shareof_newres =  round(((moved_same_county + moved_diff_county_same_state + moved_diff_state + moved_abroad) / total_pop_1yr * 100), 1)) %>%
  dplyr::select(-B07003_001M, -B07003_004M, -B07003_007M, -B07003_010M, -B07003_013M,-B07003_016M, -NAME) %>%
  st_drop_geometry()
```

## Motor Vehicles

```{r}
#| message: false
#| warning: false


philly_vehicles <- get_acs(
  geography = "tract",
  variables = c(
    "B08201_001",  # Total households
    "B08201_002"   # No vehicle available
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide",
  progress_bar = FALSE
)

# Rename and calculate share
philly_vehicles <- philly_vehicles %>%
  rename(
    total_households = B08201_001E,
    no_vehicle = B08201_002E
  ) %>%
  mutate(share_no_vehicle = round((no_vehicle / total_households * 100), 1)) %>% dplyr::select(-B08201_001M, -B08201_002M, -NAME) %>%
  st_drop_geometry()
```

## Race

```{r}
#| message: false
#| warning: false


philly_race <- get_acs(
  geography = "tract",
  variables = c(
    "B03002_001",  # Total population
    "B03002_004",  # Black alone
    "B03002_012"   # Hispanic or Latino
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide",
  progress_bar = FALSE
)

# Rename and calculate shares
philly_race <- philly_race %>%
  rename(
    total_pop = B03002_001E,
    black = B03002_004E,
    hispanic_latino = B03002_012E
  ) %>%
  mutate(
    share_black = round((black / total_pop * 100), 1),
    share_hispanic = round((hispanic_latino / total_pop * 100), 1)) %>%
  dplyr::select(-B03002_001M, -B03002_004M, -B03002_012M, -NAME) %>%
  st_drop_geometry()


```

## Rent Burden

```{r}
#| message: false
#| warning: false

philly_rent_burden <- get_acs(
  geography = "tract",
  variables = c(
    "B25070_001",  # Total renter-occupied households
    "B25070_007",  # 30.0 to 34.9 percent
    "B25070_008",  # 35.0 to 39.9 percent
    "B25070_009",  # 40.0 to 49.9 percent
    "B25070_010"   # 50.0 percent or more
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide",
  progress_bar = FALSE
)

# Rename and calculate share
philly_rent_burden <- philly_rent_burden %>%
  rename(
    total_renters = B25070_001E,
    burden_30_34 = B25070_007E,
    burden_35_39 = B25070_008E,
    burden_40_49 = B25070_009E,
    burden_50plus = B25070_010E
  ) %>%
  mutate(
    rent_burdened_30plus = burden_30_34 + burden_35_39 + burden_40_49 + burden_50plus,
    share_rent_burdened = round((rent_burdened_30plus / total_renters * 100), 1)
  ) %>%
  st_drop_geometry()


```

## Creating Panel

```{r}

#selecting what i need from eviction data 
eviction_panel<- eviction_data %>% 
  dplyr::select(GEOID, racial_majority, week, week_date, filings_2020, tract, year, month, season )

# joining the other features so far into the panel 
#landmarks
eviction_panel<- left_join(eviction_panel, landmarks, by= "tract") 

##bike
eviction_panel<- left_join(eviction_panel, bike_census, by= "tract")

eviction_panel <- eviction_panel %>%
  rename(bikeln = n)
#census variables 
eviction_panel<- left_join(eviction_panel, census, by="GEOID")

##mobility 

eviction_panel<- left_join(eviction_panel, philly_mobility, by="GEOID")

##vehicle 

eviction_panel<- left_join(eviction_panel, philly_vehicles, by="GEOID")

##race


eviction_panel<- left_join(eviction_panel, philly_race, by="GEOID")

##rent burden


eviction_panel<- left_join(eviction_panel, philly_rent_burden, by="GEOID")




```

## 311 Calls

```{r - calls_311}

help_calls <- read.csv("data/public_cases_fc.csv")
df_spatial <- help_calls[!is.na(help_calls$lon) & !is.na(help_calls$lat), ]
help_calls <- st_as_sf(df_spatial, coords = c("lon", "lat"), crs = 4326)
help_calls <- st_transform(help_calls, 2272)

# Verify
st_crs(help_calls)$epsg 

help_calls1 <- help_calls %>%
  group_by(subject) %>%
  summarise(n = n())

maintenance_complaint <-help_calls %>% 
  filter(subject == "Maintenance Complaint") 
  
maintenance_complaint <- st_transform(maintenance_complaint, st_crs(census_tracts))
maintenance_complaint <- maintenance_complaint %>% 
  st_join(census_tracts) 

maintenance_complaint<- maintenance_complaint %>% 
  rename(tract= NAMELSAD10)

maintenance_complaint_panel <- maintenance_complaint %>%
  dplyr::select(subject, tract)

maintenance_complaint_panel<- maintenance_complaint_panel %>% 
  st_drop_geometry() %>% 
  group_by(tract) %>% 
   summarise(m_complaints = n())

##panel created
eviction_panel<- left_join(eviction_panel, maintenance_complaint_panel, by= "tract")

##Replace with NA with zero
eviction_panel <- eviction_panel %>%
  dplyr::select(-ends_with("M")) %>%
  dplyr::mutate(across(c(landmarks, m_complaints, bikeln), ~replace_na(., 0))) %>%
  dplyr::select(-geometry.x)

```

## Top 100 Filers

Here we are loading in a file that the Eviction Lab has provided that lists the top 100 filers in Phialdelphia.

```{r}
top100 <- read.csv("data/philadelphia_hotspots_media_report.csv")
top100_sf <- st_as_sf(
  top100,
  coords = c("lon", "lat"),
  crs = 4326   # correct starting CRS
)

census_crs <- st_transform(census, 2272)
top100_sf <- st_transform(top100_sf, 2272)

census_crs <- census_crs %>% dplyr::select(NAME, geometry, GEOID)
##merge to eviction panel 
census_crs$hotspot_count <- lengths(
  st_intersects(census_crs, top100_sf)
)

eviction_panel<- left_join(eviction_panel, census_crs, by= "GEOID")
```

## Create Lag

```{r}

##Filter for years we want

eviction_panel_filter <- eviction_panel %>%
  filter(year(week_date) >= 2023)


eviction_panel_filter <- eviction_panel_filter %>%
  arrange(GEOID, week_date) %>%
  group_by(GEOID) %>%
  mutate(
    # Short-term lags
    filings_lag1week = lag(filings_2020, 1),
    filings_lag2week = lag(filings_2020, 2),
    filings_lag4week = lag(filings_2020, 4),

    # Medium-term lags
    filings_lag12week = lag(filings_2020, 12),
    filings_lag26week = lag(filings_2020, 26),

    # Long-term lag
    filings_lag52week = lag(filings_2020, 52),

    # Rolling average (4-week)
    roll_avg = rowMeans(
      cbind(
        lag(filings_2020, 1),
        lag(filings_2020, 2),
        lag(filings_2020, 3),
        lag(filings_2020, 4)
      ),
      na.rm = FALSE
    )
  ) %>%
  ungroup() %>%
  filter(!is.na(filings_lag52week))
  

```

## Build Model Data

Now that we have constructed the full panel dataset, we can move onto creating our predictive models.

```{r}
model_data <- eviction_panel_filter %>%
  mutate(
    season = factor(season)
  ) %>%
  dplyr::select(
    GEOID,
    tract,
    filings_2020,
    filings_lag1week,
    filings_lag2week,
    filings_lag4week,
    filings_lag12week,
    filings_lag26week,
    filings_lag52week,
    roll_avg,
    Med_Inc,
    Transit_Commuters,
    Median_age,
    shareof_newres,
    share_no_vehicle,
    share_black,
    share_hispanic,
    share_rent_burdened,
    landmarks,
    bikeln,
    m_complaints,
    month,
    season,
    week_date,
    hotspot_count,
    geometry,
    year
    
  ) %>%
  drop_na()

# converting date from character to correct date format
model_data <- model_data %>%
  mutate(week_date = as.Date(week_date))
```

## Train/Test split

```{r}
train <- model_data %>%
  filter(week_date < as.Date("2025-01-05"))

test <- model_data %>%
  filter(week_date >= as.Date("2025-01-05"))
```

## Poisson Model Baseline

```{r}

pn_model <- glm(
  filings_2020 ~  season + month, 
  data = train,
  family = poisson(link = "log")
)

summary(pn_model)

exp(coef(pn_model))

dispersion <- sum(residuals(pn_model, type = "pearson")^2)/
  pn_model$df.residual

cat("Dispersion parameter:", round(dispersion, 3), "\n")


overdisp_table <- data.frame(
  Metric = c(
    "Mean of outcome",
    "Variance of outcome",
    "Variance / Mean",
    "Dispersion (Pearson)"
  ),
  Value = c(
    mean(train$filings_2020, na.rm = TRUE),
    var(train$filings_2020, na.rm = TRUE),
    var(train$filings_2020, na.rm = TRUE) /
      mean(train$filings_2020, na.rm = TRUE),
    sum(residuals(pn_model, type = "pearson")^2) /
      pn_model$df.residual
  )
)

overdisp_table


overdisp_table.1 <- kable(
  overdisp_table,
  col.names = c("Metric", "Value"),
  caption = "Overdispersion Diagnostics",
  digits = 3,
  align = c("l", "r")
) %>%
  kable_styling(full_width = FALSE, position = "center")

overdisp_table.1

```

```{r}
nb_model <- glm.nb(
  filings_2020 ~  season + month, 
  data = train
)

summary(nb_model)

##Compare to Poisson 

AIC(nb_model)
AIC(pn_model) 

##Extract 
nb_model$theta

##Calculate MAE and RMSE
test$pred_nb <- predict(nb_model, newdata = test, type = "response")

nb_MAE <- mean(abs(test$filings_2020 - test$pred_nb))
nb_RMSE <- sqrt(mean((test$filings_2020 - test$pred_nb)^2))

nb_MAE
nb_RMSE

```

## NB + temp lag

```{r}
nb_model1 <- glm.nb(
  filings_2020 ~ season + month + roll_avg + filings_lag12week,
  data = train
)

summary(nb_model1)

nb_model1$theta

test$pred_nb1 <- predict(nb_model1, newdata = test, type = "response")

nb_MAE1 <- mean(abs(test$filings_2020 - test$pred_nb1))
nb_RMSE1 <- sqrt(mean((test$filings_2020 - test$pred_nb1)^2))

nb_MAE1
nb_RMSE1
```

## NB + demographics

```{r}

nb_model2 <- glm.nb(
  filings_2020 ~ season + roll_avg + filings_lag12week + Median_age + Med_Inc + share_black + share_hispanic,
  data = train
)

summary(nb_model2)

nb_model2$theta

test$pred_nb2 <- predict(nb_model2, newdata = test, type = "response")

nb_MAE2 <- mean(abs(test$filings_2020 - test$pred_nb2))
nb_RMSE2 <- sqrt(mean((test$filings_2020 - test$pred_nb2)^2))

nb_MAE2
nb_RMSE2


```

## NB + +rent burden, new residents and interaction effect

```{r}

nb_model3 <- glm.nb(
  filings_2020 ~ season + month + roll_avg + filings_lag12week +
    #Demographics
    Med_Inc + Median_age + share_black + share_hispanic +
    shareof_newres + 
    #Rent Burden 
    share_rent_burdened + share_black*share_rent_burdened,
  data = train
)

summary(nb_model3)

nb_model3$theta

test$pred_nb3 <- predict(nb_model3, newdata = test, type = "response")

nb_MAE3 <- mean(abs(test$filings_2020 - test$pred_nb3))
nb_RMSE3 <- sqrt(mean((test$filings_2020 - test$pred_nb3)^2))

nb_MAE3
nb_RMSE3
```

## Poisson (+landmarks, bike lanes, maintenance complaints)

```{r}
nb_model4 <- glm.nb(
  filings_2020 ~ season + month + roll_avg + filings_lag12week +
    #Demographics
    Med_Inc + Median_age + share_black + share_hispanic +
    shareof_newres + 
    #Rent Burden 
    share_rent_burdened + share_black*share_rent_burdened +
    #Landmarks, bike lanes, maintenance complaints
    landmarks + bikeln + m_complaints,
  data = train
)

summary(nb_model4)

nb_model4$theta

test$pred_nb4 <- predict(nb_model4, newdata = test, type = "response")

nb_MAE4 <- mean(abs(test$filings_2020 - test$pred_nb4))
nb_RMSE4 <- sqrt(mean((test$filings_2020 - test$pred_nb4)^2))

nb_MAE4
nb_RMSE4

```

## Poisson (+top 100 owners who have filed evictions)

```{r}
nb_model5 <- glm.nb(
   filings_2020 ~ season + month + roll_avg + filings_lag12week +
    Med_Inc + Median_age +
    shareof_newres + share_no_vehicle +
    share_black + share_hispanic + share_rent_burdened +
    share_black*share_rent_burdened + hotspot_count +
    landmarks + bikeln + m_complaints,
  data = train
)

summary(nb_model5)

nb_model5$theta


test$pred_nb5 <- predict(nb_model5, newdata = test, type = "response")

nb_MAE5 <- mean(abs(test$filings_2020 - test$pred_nb5))
nb_RMSE5 <- sqrt(mean((test$filings_2020 - test$pred_nb5)^2))

nb_MAE5
nb_RMSE5
```

## Model Evaluation

```{r - Model Evaluation}

test <- test %>%
  mutate(
    pred = predict(nb_model,  newdata = test, type = "response"),
    pred1 = predict(nb_model1, newdata = test, type = "response"),
    pred2 = predict(nb_model2, newdata = test, type = "response"),
    pred3 = predict(nb_model3, newdata = test, type = "response"),
    pred4 = predict(nb_model4, newdata = test, type = "response"),
    pred5 = predict(nb_model5, newdata = test, type = "response")
  )

mae_baseline <- data.frame(
  Model = c("1. Baseline season",
            "2. + temporal lags",
            "3. + demographics",
            "4. + rent burden, mobility, interaction effect",
            "5. + landmarks, bike lanes, and maintenance complaints",
            "6. + top 100 property owners filing evictions"),
  MAE   = c(
    mean(abs(test$filings_2020 - test$pred), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred1), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred2), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred3), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred4), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred5), na.rm = TRUE)
  )
)

kable(mae_baseline, digits = 3,
      caption = "MAE by model — 2025 test set",
      col.names = c("Model", "MAE")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

model_comparison <- ggplot(mae_baseline, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = .8) +
  geom_text(aes(label = round(MAE, 3)), vjust = -0.5) +
  labs(title = "Model Performance Comparison",
       x = "Model", y = "MAE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1)  # Start at 1, auto-calculate upper limit

model_comparison
```

The temporal lags are doing all the work here. Adding predictors only reduced MAE by a very small amount. That being said it is still a reasonable MAE on the whole.

# Space-time error analysis (Model 5)

## Observed vs. predicted

This model is more conducive to a diagnostics plot for our logistic regression. It works by first separating the predicted (model5) probabilities into 10 equal sized bins so 0.0-01...

Hence all tract-weeks are sorted based on their predictive probability. Within each bin two values are calculated. First the mean predictive probability. This looks at all tract-weeks in the first bin and finds the average predictive probability. So say there are 400 tract-weeks in the bin where our model5 predicted the chance of eviction is between 0 and 10%. It looks at the distribution of probabilities within this bin and averages it for us. In this case say the average predictive probability is 7%. So 7% of our 400 observations should have an eviction filing (28). We can then actually compare this to the real value which is the mean of whether an eviction occurred or not. We would expect a corresponding 28/400 real eviction filings within this bin. We can then plot these against each other for every bin. A straight line 45 degrees means our model is predicting well.

## Spatial errors

```{r - Spatial Errors}
#| message: false
#| warning: false

# --- Create prediction error ---
test <- test %>%
  mutate(
    error = filings_2020 - pred5,
    abs_error = abs(error),
    pct_error = (pred5 - filings_2020) / (filings_2020 +1) * 100)

# --- Aggregate errors at tract level ---
tract_errors <- test %>%
  filter(!is.na(pred5)) %>%
  group_by(GEOID) %>%
  summarise(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_eviction_filing = mean(filings_2020, na.rm = TRUE),
    prediction = mean(pred5, na.rm = TRUE),
    .groups = "drop"
  )

# --- Join to spatial geometry ---
tract_errors_sf <- census_tracts %>%
  left_join(tract_errors, by = "GEOID")

# --- Map 1: Prediction error ---
p1 <- ggplot() +
  geom_sf(
    data = tract_errors_sf,
    aes(fill = MAE),
    color = "white",
    linewidth = 0.1
  ) +
   # --- North arrow ---
  annotation_north_arrow(
    location = "br",   # top right
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # --- Scale bar ---
  annotation_scale(
    location = "bl",   # bottom left
    width_hint = 0.3
  ) +
    scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Prediction Error (MAE)") +
 theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

# --- Map 2: Observed eviction rates ---
p2 <- ggplot() +
  geom_sf(
    data = tract_errors_sf,
    aes(fill = avg_eviction_filing),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Observed Eviction Filing") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))


p3 <- ggplot() +
  geom_sf(
    data = tract_errors_sf,
    aes(fill = prediction),
    color = "white",
    linewidth = 0.1
  ) +
  scale_fill_viridis_c(option = "plasma", direction = -1)+
  labs(title = "Predicted Eviction Filing") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))


(p1)

(p2)

(p3)


```

```{r - Temporal Errors}
#| message: false
#| warning: false

# Aggregate Model 5 errors by season and pre/post-law period
temporal_errors <- test %>%
  filter(!is.na(pred5)) %>%
  mutate(
    season         = factor(season, levels = c("winter", "spring", "summer", "fall"))
  ) %>%
  group_by(season) %>%
  summarise(
    MAE = mean(abs_error, na.rm = TRUE),
    n   = n(),
    .groups = "drop"
  )

p9 <- ggplot(temporal_errors, aes(x = season, y = MAE,)) +
  geom_col(fill = "#3182bd", position = "dodge") +
  geom_text(aes(label = round(MAE, 2)), vjust = -.5) +
  labs(
    title = "Prediction errors by season",
    x     = "Season",
    y     = "MAE",
    fill  = "Policy period"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

```{r - Demographic Errors}

# Aggregate Model 5 errors and demographics by tract
tract_errors_demo <- test %>%
  filter(!is.na(pred5)) %>%
  group_by(GEOID) %>%
  summarise(
    MAE                  = mean(abs_error, na.rm = TRUE),
    Med_Inc              = first(Med_Inc),
    share_rent_burdened  = first(share_rent_burdened),
    share_black          = first(share_black),
    share_hispanic       = first(share_hispanic),
    .groups              = "drop"
  ) %>%
  filter(!is.na(Med_Inc))

p4 <- ggplot(tract_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. median income",
       x = "Median household income", y = "MAE") +
  theme_minimal()

p5 <- ggplot(tract_errors_demo, aes(x = share_rent_burdened, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. rent burden",
       x = "% rent-burdened (30%+ of income on rent)", y = "MAE") +
  theme_minimal()

p6 <- ggplot(tract_errors_demo, aes(x = share_black, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. share Black",
       x = "% Black", y = "MAE") +
  theme_minimal()

p7 <- ggplot(tract_errors_demo, aes(x = share_hispanic, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. share Hispanic",
       x = "% Hispanic", y = "MAE") +
  theme_minimal()

(p1 | p2) / (p3 | p4)
p1
p2
p3
p4
```

Only real large relationship is between MAE Errors and Median Household Income. Model5 is far better at predicting the liklihood of eviction filings in census tracts with high median household income.

```{r}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error     = filings_2020 - pred5,
    abs_error = abs(error),
    pct_error = (pred5 - filings_2020)/(filings_2020 +1) *100
  )


##Visualize
p10 <- ggplot(test, aes(x = filings_2020, y = pred5)) +
  geom_jitter(alpha = 0.2, color = "#3182bd", width = 0.05, height = 0) +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(season ~ year) +
  labs(title = "Observed vs. predicted — 2025 test set",
       subtitle = "Model 5; red = perfect calibration, green = fitted trend",
       x = "Observed (1 = eviction filed)", y = "Predicted probability") +
  theme_minimal()



```

## Comparison Results

```{r}
# create table of comparison rmse, mae
model_names <- c("1. Baseline season", "2. + temporal lags", "3. + demographics", "4. + rent burden, mobility, interaction effect", "5. + landmarks, bike lanes, and maintenance complaints", "6. + top 100 property owners filing evictions (Final)")

models <- list(nb_model, nb_model1, nb_model2, nb_model3, nb_model4, nb_model5)

comparison <- data.frame(
  Model = model_names,
  MAE = round(c(nb_MAE, nb_MAE1, nb_MAE2, nb_MAE3, nb_MAE4, nb_MAE5), 2),
  RMSE = round(c(nb_RMSE, nb_RMSE1, nb_RMSE2, nb_RMSE3, nb_RMSE4, nb_RMSE5), 2)
)

comparison_m <- kable(comparison,
      col.names = c("Model", "MAE", "RMSE"),
      align = "lccc",
      caption = "Each layer prediction improves prediction") %>%
  kable_styling(font_size = 20, full_width = FALSE) %>%
  row_spec(nrow(comparison), bold = TRUE, background = "#27AE60")

comparison_m
```

```{r}
##Maps rent burden
library(ggspatial)

philly_rent_burden1 <- left_join(census_crs, philly_rent_burden,  by = "GEOID")

rent_map <- ggplot(philly_rent_burden1) +
  geom_sf(aes(fill = share_rent_burdened), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c() +
  
  # --- North arrow ---
  annotation_north_arrow(
    location = "br",   # top right
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # --- Scale bar ---
  annotation_scale(
    location = "bl",   # bottom left
    width_hint = 0.3
  ) +
  
  labs(
    title = "Share of People who are Rent Burdened by Census Tract",
    fill = "Share of People Whoe Are Rent Burdened"
  ) +
  theme_void()

rent_map

philly_race1 <- left_join(census_crs, philly_race,  by = "GEOID")

black <- ggplot(philly_race1) +
  geom_sf(aes(fill = share_black), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c() +
  
  # --- North arrow ---
  annotation_north_arrow(
    location = "br",   # top right
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # --- Scale bar ---
  annotation_scale(
    location = "bl",   # bottom left
    width_hint = 0.3
  ) +
  
  labs(
    title = "Share of Black Population by Census Tract",
    fill = "Share of Black Population"
  ) +
  theme_void()

black

hispanic <- ggplot(philly_race1) +
  geom_sf(aes(fill = share_hispanic), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c() +
  
  # --- North arrow ---
  annotation_north_arrow(
    location = "br",   # top right
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # --- Scale bar ---
  annotation_scale(
    location = "bl",   # bottom left
    width_hint = 0.3
  ) +
  
  labs(
    title = "Share of Hispanic Population by Census Tract",
    fill = "Share of Hispanic Population"
  ) +
  theme_void()

hispanic



```

```{r}
save(eviction_map, p1, p2, p3, p4, p5, p6, p7, p9, p10, seasonal_patterns, weekly_plot, mae_baseline, model_comparison, overdisp_table.1, eviction_map, comparison_m, rent_map, black, hispanic, file = "final_project.RData")
```
