---
title: "Predicting Eviction Filings in Philadelphia"
subtitle: "A Census-Tract Level Negative Binomial Model"
author: "Huda, Dalanda, Wesley, Yusuf"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
---



```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(sf)
library(tmap)
library(ggplot2)
library(kableExtra)
library(viridis)
library(patchwork)
library(pROC)

load("final_project.RData")

```

## Problem & Policy Context

Extent literature shows the devastating effects of evictions on health, social, and economic outcomes. Evictions are spatially correlated and persistent over time.

A targeted, more focused approach to eviction prevention is important in Philadelphia given that it is the second poorest city in the U.S. There were over **13,000** eviction filings in the past 12 months in the city.

Black & female Philadelphians were more likely to receive an eviction threat in the last 12 months than their white and male counterparts, respectively.

## Motivation

We create a predictive model for eviction filings to identify areas wherein to target eviction prevention strategies and resources more effectively.

## Research Question

Which census tracts in Philadelphia are most at risk of receiving eviction filings?

Which variables best predict eviction filings?

## Data Sources

- Eviction filings data, 2020–2026 (Eviction Lab)
- Census ACS (ACS5, 2022) — income, race, mobility, rent burden, vehicle access
- OpenDataPhilly — bike network, landmarks, 311 maintenance complaints
- Eviction Lab Top 100 Filers (Philadelphia hotspots)
- Census tract boundaries (TIGER/Line)
- Our integration approach was joining our data by census tract

## Demographic Overview
```{r}
rent_map
```

## Demographic Overview
```{r}
black
```


## Demographic Overview
```{r}
hispanic
```


## Where Are Evictions Concentrated?
```{r}
eviction_map
```

## Temporal Patterns

```{r}
seasonal_plot <- ggplot(seasonal_patterns, aes(x = season, y = avg_filings, group = 1)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Seasonal Eviction Patterns",
    x = "Season",
    y = "Average Filings"
  )
seasonal_plot 


```


## Temporal Patterns
```{r}
weekly_plot
```

## Methods Overview
:::: {.columns}
::: {.column width="50%"}

**Negative Binomial**

- Count data
- Poisson: overdispersion
- Logistic: lose too much detail


**Integration Approach**

- Joined using GEOID / Census Tract
- Unit of analysis: filings by week, month, season

:::
::::{.column width="50%"}

```{r}
overdisp_table.1
```

:::
::::

## Model Building

*We built 5 models in sequence*

- Model 1: baseline (season + month)
- Model 2: + temporal lags
- Model 3: + demographics (age, race, income)
- Model 4: + rent burden, new residents, transit commuters
- Model 5: + physical features (landmarks, bike lanes, maintenance complaints)
- Model 6: + top 100 owners who filed evictions in Philadelphia

## Model Performance

```{r}

model_comparison
```

## Spatial Errors

```{r}
p1 + 
p2 +
p3
```



## Error vs Demographics 
```{r}
p4+
p5+
p6+
p7
```


## Error vs Season 

```{r}
p9
```



## Predicted vs Observed

```{r}
p10
```

## Final Model Comparison
```{r}
comparison_m
```


## Recommendations

- Promote intervention program that supports census tracts that have high counts of eviction filings
- Conduct outreach and assessment in tracts with high counts of eviction filings
- Connect non-profit community organizations with identified census tracts to allocate resources
- Investigate top 100 eviction filers

## Limitations and Bias Considerations

- Data is aggregated at the tract level, so we miss intra-tract differences
- Data captures filings not processed evictions
- Eviction rates are higher for Black renters than white renters
- Historical inequities shape socioeconomic qualities like income and rent burden
    - This type of model could be misused to stigmatize poorer and Black neighborhoods
    
## Questions?

Thank you
