---
title: "Lab 0: Getting Started with dplyr"
subtitle: "Your First Data Analysis"
author: "Wesley Nay"
date: 01-20-2026
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Overview

Welcome to your first lab! In this (not graded) assignment, you'll practice the fundamental dplyr operations I overviewed in class using car sales data. This lab will help you get comfortable with:

- Basic data exploration
- Column selection and manipulation  
- Creating new variables
- Filtering data
- Grouping and summarizing

**Instructions:** Copy this template into your portfolio repository under a `lab_0/` folder, then complete each section with your code and answers. You will write the code under the comment section in each chunk.
Be sure to also copy the data folder into your `lab_0` folder.

# Setup

```{r setup}
# Load the tidyverse library
library(tidyverse)

# Read in the car sales data
# Make sure the data file is in your lab_0/data/ folder
car_data <- read_csv("data/car_sales_data.csv")
```

# Exercise 1: Getting to Know Your Data

## 1.1 Data Structure Exploration

Explore the structure of your data and answer these questions:

```{r explore-structure}
# Use glimpse() to see the data structure
glimpse(car_data)

# Check the column names
colnames(car_data)

# Look at the first few rows
head(car_data)
```

**Questions to answer:**
- How many rows and columns does the dataset have?

- What types of variables do you see (numeric, character, etc.)?

- Are there any column names that might cause problems? Why?


**Your answers:**
- Rows: 50,000
- Columns: 7  
- Variable types: character and double (numeric)
- Problematic names: 'Engine size', 'Fuel type', 'Year of manufacture' because they have spaces and cannot be referenced in functions. To reference these columns, you have to use quotation marks.

## 1.2 Tibble vs Data Frame

Compare how tibbles and data frames display:

```{r tibble-vs-dataframe}
# Look at the tibble version (what we have)
car_data

# Convert to regular data frame and display
car_df <- as.data.frame(car_data)
car_df
```

**Question:** What differences do you notice in how they print?

**Your answer:** I don't see any differences in how they print inline with the script besides the description indicating tibble or dataframe (df).

# Exercise 2: Basic Column Operations

## 2.1 Selecting Columns

Practice selecting different combinations of columns:

```{r column-selection}
# Select just Model and Mileage columns
select(car_data, Model, Mileage)

# Select Manufacturer, Price, and Fuel type
select(car_data, Manufacturer, Price, "Fuel type")


# Challenge: Select all columns EXCEPT Engine Size
select(car_data, -"Engine size")

```

## 2.2 Renaming Columns

Let's fix a problematic column name:

```{r rename-columns}
# Rename 'Year of manufacture' to year
car_data <- rename(car_data, year = "Year of manufacture")

# Check that it worked
names(car_data)
```

**Question:** Why did we need backticks around `Year of manufacture` but not around `year`?

**Your answer:** Because there are spaces between 'Year of manufacture', and the backticks identify the three words as one variable in R.

# Exercise 3: Creating New Columns

## 3.1 Calculate Car Age

```{r calculate-age}
# Create an 'age' column (2025 minus year of manufacture)
car_data <- mutate(car_data, age = 2025 - year)

# Create a mileage_per_year column  
car_data <- mutate(car_data, mileage_per_year = Mileage / age)

# Look at your new columns
select(car_data, Model, year, age, Mileage, mileage_per_year)
```

## 3.2 Categorize Cars

```{r categorize-cars}
# Create a price_category column where if price is < 15000, it is coded as budget, between 15000 and 30000 is midrange and greater than 30000 is high-end (use case_when)
car_data <- mutate(car_data, 
                   price_category = case_when(
                     Price < 15000 ~ "budget",
                     Price > 15000 & Price < 30000 ~ "midrange",
                     Price > 30000 ~ "high-end"))

# Check your categories select the new column and show it
select(car_data, Price, price_category)
```

# Exercise 4: Filtering Practice

## 4.1 Basic Filtering

```{r basic-filtering}
# Find all Toyota cars
toyotas <- filter(car_data, Manufacturer == "Toyota")

# Find cars with mileage less than 30,000
low_mileage <- filter(car_data, Mileage < 30000)

# Find luxury cars (from price category) with low mileage
lux_lowmil <- filter(low_mileage, price_category == "high-end")

```

## 4.2 Multiple Conditions

```{r multiple-conditions}
# Find cars that are EITHER Honda OR Nissan
honda_nissan <- filter(car_data, Manufacturer %in% c("Honda", "Nissan"))
#unique(car_data$Manufacturer)
##There are no Hondas or Nissans, only Ford, Prosche, Toyota, VW, and BMW

# Find cars with price between $20,000 and $35,000
price_filtered_20to35 <- filter(car_data, Price %in% (20000:35000))

# Find diesel cars less than 10 years old
car_data <- car_data %>%
  rename(fuel_type = "Fuel type")

diesel <- filter(car_data, fuel_type == "Diesel")

new_diesel <- filter(diesel, age < 10)

```

**Question:** How many diesel cars are less than 10 years old?

**Your answer:** 2,040

# Exercise 5: Grouping and Summarizing

## 5.1 Basic Summaries

```{r basic-summaries}
# Calculate average price by manufacturer
avg_price_by_brand <- car_data %>%
  group_by(Manufacturer) %>%
  summarize(avg_price = mean(Price, na.rm = TRUE))

avg_price_by_brand

# Calculate average mileage by fuel type
avg_mileage_by_fuel_type <- car_data %>%
  group_by(fuel_type) %>%
  summarise(avg_mileage = mean(Mileage, na.rm = TRUE))

avg_mileage_by_fuel_type

# Count cars by manufacturer
cars_by_manuf <- car_data %>%
  count(Manufacturer)

cars_by_manuf

```

## 5.2 Categorical Summaries

```{r categorical-summaries}
# Frequency table for price categories
freq_price_cat <- car_data %>%
  count(price_category)

print(freq_price_cat)
```


# Submission Notes

**To submit this lab:**
1. Make sure your code runs without errors
2. Fill in all the "[YOUR ANSWER]" sections and complete all of the empty code! 
3. Save this file in your portfolio's `lab_0/` folder
4. Commit and push to GitHub
5. Check that it appears on your GitHub Pages portfolio site

**Questions?** Post on the canvas discussion board or come to office hours!