---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Wesley Nay"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Illinois Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("f767d3579b2eac0cc1f4a249267bd88ef4765608")
# Choose your state for analysis - assign it to a variable called my_state
my_state <- "IL"

```

**State Selection:** I have chosen **Illinois** for this analysis because: it is my home state.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
il_data <- get_acs(
  geography = "county",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001"
  ),
  state = my_state,
  year = 2022,
  output = "wide"
)

# Clean the county names to remove state name and "County" 
il_clean <- il_data %>%
  mutate(
    # Remove state name from county names
    county_name = str_remove(NAME, ", Illinois"),
    # Remove "County" word
    county_name = str_remove(county_name, " County")
  )

# Hint: use mutate() with str_remove()

# Display the first few rows
head(il_clean)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
il_reliability <- il_clean %>%
  mutate(
    # Calculate MOE as percentage of estimate
    moe_percentage = round((median_incomeM / median_incomeE) * 100, 2),
     # Create reliability categories
    reliability = case_when(
      moe_percentage < 5 ~ "High Confidence",
      moe_percentage >= 5 & moe_percentage <= 10 ~ "Moderate Confidence",
      moe_percentage > 10 ~ "Low Confidence"
    )
  )

# Create a flag for unreliable estimates (MOE > 10%)
il_reliability <- il_reliability %>%
  mutate(unreliable = moe_percentage > 10)

# Create a summary showing count of counties in each reliability category
il_reliability_summary <- il_reliability %>%
  count(reliability)

# Hint: use count() and mutate() to add percentages
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
high_uncertainty <- il_reliability %>%
  arrange(desc(moe_percentage)) %>%
  select(
    county_name,
    median_incomeE,
    median_incomeM,
    moe_percentage,
    reliability
  ) %>%
  slice(1:5)


# Format as table with kable() - include appropriate column names and caption
kable(
  high_uncertainty,
  col.names = c(
    "County",
    "Median Income",
    "Margin of Error",
    "MOE (%)",
    "Reliability Category"
  ),
  caption = "Counties with Highest Income Data Uncertainty",
  format.args = list(big.mark = ",")
)
```

**Data Quality Commentary:**

These counties are at risk of being poorly served by algorithms that rely on median household income as these estimates may not accurately reflect local economic conditions. Factors that might contribute to higher uncertainty in this case include small population size, limited survey responses, and greater variance of median household income among the surveyed respondents.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
# Select one county from each reliability category
selected_counties <- il_reliability %>%
  filter(
    county_name %in% c("Cook", "Carroll", "Pulaski")
  )

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties %>%
  select(
    county_name,
    median_incomeE,
    moe_percentage,
    reliability
  )


```

**Comment on the output:** I selected Cook County (high confidence), Carroll County (moderate confidence), and Pulaski County (low confidence). Cook County contains the Chicago Metropolitan area and has the highest median income among the three selected counties. Cook County has about 5.2 million total residents, while Carroll and Pulaski have about 15,000 and 5,000, respectively. The significant difference in population totals can explain the variance in margin of errors.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
race_vars <- c(
  total_pop = "B03002_001",
  white = "B03002_003",
  black = "B03002_004",
  hispanic = "B03002_012"
)

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
selected_county_fips <- selected_counties %>%
  mutate(county_fips = substr(GEOID, 3, 5)) %>%
  pull(county_fips)

# Retrieve tract-level demographic data
tract_data <- get_acs(
  geography = "tract",
  variables = race_vars,
  state = my_state,
  county = selected_county_fips,
  year = 2022,
  survey = "acs5",
  output = "wide"
)

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations

tract_demo <- tract_data %>%
  mutate(
    pct_white = (whiteE / total_popE) * 100,
    pct_black = (blackE / total_popE) * 100,
    pct_hispanic = (hispanicE / total_popE) * 100,
    county_fips = substr(GEOID, 3, 5),
    tract_id = substr(GEOID, 6, 11)
  )

# Add readable tract and county name columns using str_extract() or similar
tract_demo <- tract_demo %>%
  mutate(
    tract_name = str_extract(NAME, "Census Tract [^;]+"),       # Census Tract ID
    county_name = str_extract(NAME, "[^;]+ County"),            # "Carroll County"
    county_name = str_remove(county_name, " County")            # remove " County" for matching
  )


```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
highest_hispanic_tract <- tract_demo %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(
    tract_name,
    county_name,
    pct_hispanic,
    pct_white,
    pct_black,
    total_popE
  )


# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
county_demographics <- tract_demo %>%
  group_by(county_name) %>%
  summarize(
    number_of_tracts = n(),                         # counts tracts per county
    avg_pct_white = mean(pct_white, na.rm = TRUE),  # average % white
    avg_pct_black = mean(pct_black, na.rm = TRUE),  # average % black
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE) # average % Hispanic
  ) %>%
  ungroup()  # remove grouping

# Create a nicely formatted table of your results using kable()
kable(
  county_demographics,
  digits = 1,
  col.names = c(
    "County",
    "Number of Tracts",
    "Avg % White",
    "Avg % Black",
    "Avg % Hispanic/Latino"
  ),
  caption = "Average Tract-Level Demographics by County"
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
demographic_moe <- tract_demo %>%
  mutate(
    # Calculate MOE as percentage of estimate
    moe_pct_white = round((whiteM / whiteE) * 100, 2),
    moe_pct_black = round((blackM / blackE) * 100, 2),
    moe_pct_hispanic = round((hispanicM / hispanicE) * 100, 2),
    # Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
    # Flag tracts where any demographic MOE > 15%
    high_moe_flag = ifelse(
      moe_pct_white > 15 | moe_pct_black > 15 | moe_pct_hispanic > 15,
      TRUE,
      FALSE
    )
  )

# Create summary statistics showing how many tracts have data quality issues
demographic_moe_summary <- demographic_moe %>%
  summarize(
    total_tracts = n(),
    tracts_high_moe = sum(high_moe_flag),
    pct_high_moe = round((tracts_high_moe / total_tracts) * 100, 1)
  )

# Display summary
demographic_moe_summary
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
pattern_analysis <- demographic_moe %>%
  mutate(high_moe_flag = ifelse(high_moe_flag, "High MOE", "Low MOE"))  # convert to descriptive label

# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison
pattern_analysis <- pattern_analysis %>%
  group_by(high_moe_flag) %>%
  summarize(
    number_of_tracts = n(),
    avg_total_pop = round(mean(total_popE, na.rm = TRUE), 0),
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 1),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 1),
    avg_pct_hispanic = round(mean(pct_hispanic, na.rm = TRUE), 1)
  )

# Create a professional table showing the patterns
kable(
  pattern_analysis,
  col.names = c(
    "MOE Category",
    "Number of Tracts",
    "Avg Total Population",
    "Avg % White",
    "Avg % Black",
    "Avg % Hispanic/Latino"
  ),
  caption = "Comparison of Tracts with High vs Low MOE"
)

```

**Pattern Analysis:**

Based on these outputs, Census Tracts with smaller populations tend to have less reliable data. Additionally, tracts that appear to be more racially/ethnically diverse have higher MOEs (less reliable data). Given historic discriminatory patterns by the U.S. government against Black and Latino/Hispanic populations, we may infer that these tracts may have lower survey response rates leading to less reliable data. These communities may not feel comfortable sharing personal information with the U.S. government.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**
***Overall Pattern Identification***
Across my analyses, I find that data reliability varies systematically by geography and population characteristics like median income and race. At the county level in Illinois, about a third of the ACS estimates for median income are reliable, a few more counties are moderately reliable, and a few smaller and/or rural counties are less reliable. At the tract level, margins of error for racial/ethnic demographic variables are concentrated in smaller, more diverse communities. This indicates that data uncertainty is not randomly distributed. These patterns suggest that both income and demographic data may misrepresent certain communities if used uncritically in algorithmic decision-making.

***Equity Assessment***
The communities most at risk off algorithmic bias are small and/or raciall/ethnically diverse census tracts. For example tracts with high margins of error often have lower percentages of white residents and higher percentages of Black and Hispanic/Latino residents. High margin of error tracts represent areas where demographic characteristics are less reliably captured, which could lead to systematic underrepresentation in funding and outreach decisions.

***Root Cause Analysis***
The underlying factors that drive both data quality issues and bias risk are small population sizes, limited survey responses, and income variability within small geographic areas. Small population sizes can be problematic in survey data because surveys have to create estimates based on responses. So, in areas with fewer people, these estimates are less reliable because there are not enough survey responses to give a stable number. These factors contribute both to higher margins of error and to potential bias in algorithmic outputs, creating systematic risks for certain populations.

***Strategic Recommendations***
To address these challenges, the Department should adopt a tiered approach to algorithmic decision-making. Counties and tracts with high-confidence data can be used directly to guide algorithmic resource allocation. Moderate-confidence areas should be monitored closely, with adjustments made as more reliable data becomes available. Low-confidence areas should be supplemented with manual review, local surveys, and/or alternative data sources before making any automated decisions. Additionally, algorithmic models should include data reliability/uncertainty metrics to weight decisions.


## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

algorithm_recommendations <- il_reliability %>%
  select(county_name, median_incomeE, moe_percentage, reliability)


# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"
algorithm_recommendations <- algorithm_recommendations %>%
  mutate(
  algorithm_recommendation = case_when(
    reliability == "High Confidence" ~ "Safe for algorithmic decisions",
    reliability == "Moderate Confidence" ~ "Use with caution - monitor outcomes",
    reliability == "Low Confidence" ~ "Requires manual review or additional data"
  )
)

# Format as a professional table with kable()
kable(
  algorithm_recommendations,
  col.names = c(
    "County",
    "Median Income",
    "MOE (%)",
    "Reliability Category",
    "Algorithm Recommendation"
  ),
  caption = "County-Level Algorithm Implementation Recommendations",
  format.args = list(big.mark = ",")
)

```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:**
*Adams, Boone, Champaign, Clinton, Cook, DeKalb, DuPage, Fulton, Greene, Grundy, Jefferson, Kane, Kankakee, Kendall, Lake, LaSalle, Lee, Livingston, Logan, McHenry, McLean, Macon, Macoupin, Madison, Marion, Mercer, Monroe, Ogle, Peoria, Rock Island, St. Clair, Sangamon, Shelby, Stephenson, Tazewell, Vermilion, Will, Winnebago*

The counties listed above are suitable for immediate algorithmic implementation because they have high confidence data (the margin of error is less than 5%), which means that median household income and demographic estimates are reliable. Algorithmic decisions based on these data are unlikely to misrepresent communities.

2. **Counties requiring additional oversight:**
*Bond, Bureau, Calhoun, Carroll, Cass, Christian, Clark, Clay, Coles, Crawford, De Witt, Douglas, Edwards, Effingham, Fayette, Ford, Franklin, Hancock, Henderson, Henry, Iroquois, Jackson, Jasper, Jersey, Jo Daviess, Johnson, Knox, McDonough, Marshall, Mason, Massac, Menard, Montgomery, Morgan, Perry, Pike, Putnam, Randolph, Richland, Saline, Scott, Stark, Union, Washington, Wayne, White, Whiteside, Williamson, Woodford*

The counties listed above have moderate confidence data (margin of error between 5 and 10%), so algorithms *may* be used here **cautiously** with additional oversight. If algorithms are used in decision-making here, outputs *must* be monitored and validated periodically. There should be key performance indicators that are tracked over time. Critical decisions should be supplemented with local knowledge or updated survey data to ensure equity. For example, if there are more detailed and accurate records/data, algorithmic outputs should be cross-checked and validated against these records.

3. **Counties needing alternative approaches:**

*Alexander, Brown, Cumberland, Edgar, Gallatin, Hamilton, Hardin, Lawrence, Moultrie, Piatt, Pope, Pulaski, Schuyler, Wabash, Warren*

The counties listed above have low-confidence data (margin of error is greater than 10%) and require alternative approaches to decision-making. Alternative approaches include manual review, additional local surveys, and combining multiple data sources to improve reliability before implementing any algorithmic allocation.

## Questions for Further Investigation

1. Are high-MOE tracts concentrated in certain geographic regions or types of communities (e.g., rural vs urban)?
2. How does place of birth affect MOE? What about primary language spoken at home? Highest level of education?
3. How does MOE vary by area based on historic redlining maps?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on 2/9/2026

**Reproducibility:** 
- All analysis conducted in R version 4.5.2 (2025-10-31 ucrt)
- Census API key required for replication
- Complete code and documentation available at: https://wesleynay.github.io/cpln5920/

**Methodology Notes:**
County selection for Part 3 was based on reliability categories to represent high, moderate, and low-confidence areas.
High confidence: MOE < 5%
Moderate connfidence: MOE <= 10%
Low confidence: MOE > 10%

Data cleaning involved removing state and “County” from county names and creating tract-level identifiers for easier aggregation.

**Limitations:**
ACS data are estimates with inherent sampling variability, especially in small or minority-dense tracts.

Analysis is limited to Illinois; results may not generalize to other states.

Temporal factors: data reflect 2022 estimates, so economic or demographic changes after this year are not captured.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [X] All code chunks run without errors
- [X] All "[Fill this in]" prompts have been completed
- [X] Tables are properly formatted and readable
- [X] Executive summary addresses all four required components
- [X] Portfolio navigation includes this assignment
- [X] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`